---
title: "Processing Drug-Related ICD-10 Codes in Claims Data"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Processing Drug-Related ICD-10 Codes in Claims Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to use the `process_drug_icd10_vectorized()` function to identify drug-related diagnoses in healthcare claims data. The function creates binary indicators for various substance use disorders, poisonings, and related conditions based on ICD-10-CM codes.

## Getting Started

First, load the DrugDataCoding package:

```{r}
library(DrugDataCoding)
```

## Sample Data

Let's create a sample dataset containing ICD-10 diagnosis codes:

```{r}
# Create a sample claims dataset
sample_claims <- data.frame(
  patient_id = 1:5,
  I10_DX1 = c("F11.20", "F10.10", "G89.29", "F43.21", "T40.1X1A"),
  I10_DX2 = c("Z79.891", "K70.0", NA, "F32.9", "X42"),
  I10_DX3 = c("T40.1X1A", "T51.0X1A", NA, NA, "F11.20"),
  I10_DX4 = c(NA, NA, NA, NA, NA),
  stringsAsFactors = FALSE
)

sample_claims
```

## Processing the Data

Now we'll use the `process_drug_icd10_vectorized()` function to identify drug-related diagnoses:

```{r}
processed_claims <- process_drug_icd10_vectorized(sample_claims)
```

Let's examine the results:

```{r}
# Look at the new indicators added to the dataset
names(processed_claims)[-(1:5)]  # Show only the new variables

# Show a subset of the indicators for each patient
result_subset <- processed_claims[, c("patient_id", "opioid_use_disorder", 
                                     "heroin_poisoning", "alcohol_use_disorder", 
                                     "alcohol_poisoning", "any_drug_use_disorder", 
                                     "any_drug_poisoning")]
result_subset
```

## Understanding the Results

Let's go through each patient in our sample dataset:

1. **Patient 1**: Has codes F11.20 (Opioid dependence, uncomplicated), Z79.891 (Long-term use of opiate analgesic), and T40.1X1A (Poisoning by heroin, accidental, initial encounter)
   - Identified as having an opioid use disorder and heroin poisoning

2. **Patient 2**: Has codes F10.10 (Alcohol abuse, uncomplicated), K70.0 (Alcoholic fatty liver), and T51.0X1A (Toxic effect of ethanol, accidental, initial encounter)
   - Identified as having an alcohol use disorder and alcohol poisoning

3. **Patient 3**: Has code G89.29 (Other chronic pain)
   - No drug-related indicators identified

4. **Patient 4**: Has codes F43.21 (Adjustment disorder with depressed mood) and F32.9 (Major depressive disorder, single episode, unspecified)
   - No drug-related indicators identified

5. **Patient 5**: Has codes T40.1X1A (Poisoning by heroin, accidental, initial encounter), X42 (Accidental poisoning by narcotics), and F11.20 (Opioid dependence, uncomplicated)
   - Identified as having an opioid use disorder and heroin poisoning

## Using Custom Variable Names

If your dataset uses different variable names for the diagnosis fields, you can specify them using the `dx_vars` parameter:

```{r}
# Create a sample with custom diagnosis field names
sample_custom <- data.frame(
  patient_id = 1:3,
  DX_1 = c("F11.20", "F10.10", "F12.10"),
  DX_2 = c("T40.1X1A", "K70.0", "F32.9"),
  DX_3 = c(NA, NA, NA),
  stringsAsFactors = FALSE
)

# Process using custom variable names
processed_custom <- process_drug_icd10_vectorized(sample_custom, 
                                                 dx_vars = c("DX_1", "DX_2", "DX_3"))

# Check results
processed_custom[, c("patient_id", "opioid_use_disorder", "heroin_poisoning", 
                    "alcohol_use_disorder", "cannabis_use_disorder")]
```

## Summary

The `process_drug_icd10_vectorized()` function allows you to efficiently identify a comprehensive set of drug-related conditions in healthcare claims data. This can be valuable for:

- Epidemiological studies of substance use disorders
- Health services research focusing on drug-related healthcare utilization
- Quality improvement initiatives targeting substance use care
- Surveillance of drug poisonings and overdoses

The function handles large datasets efficiently through vectorized operations and can be adapted to different data structures through the `dx_vars` parameter.
