---
title: "Processing Drug-Related ICD-10 Codes in Mortality Data"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Processing Drug-Related ICD-10 Codes in Mortality Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to use the `process_death_records_extended()` function to identify drug-related deaths in mortality data. The function creates binary indicators for drug overdose deaths and specific drug poisonings based on ICD-10 codes, following CDC definitions.

## Getting Started

First, load the DrugDataCoding package:

```{r}
library(DrugDataCoding)
```

## Sample Data

Let's create a sample dataset containing mortality records with ICD-10 cause of death codes:

```{r}
# Create a sample mortality dataset
sample_mortality <- data.frame(
  record_id = 1:5,
  UCOD = c("X42", "X44", "I21.9", "Y14", "V89.2"),  # Underlying cause of death
  MCOD1 = c("T40.1", "T40.4", "I10", "T40.2", "S06.9"),  # Multiple causes of death
  MCOD2 = c("F11.2", "T43.6", "E11.9", NA, "S27.0"),
  MCOD3 = c(NA, NA, NA, NA, NA),
  stringsAsFactors = FALSE
)

sample_mortality
```

## Processing the Data

Now we'll use the `process_death_records_extended()` function to identify drug-related deaths:

```{r}
processed_mortality <- process_death_records_extended(sample_mortality)
```

Let's examine the results:

```{r}
# Look at the new indicators added to the dataset
names(processed_mortality)[-(1:5)]  # Show only the new variables

# Show a subset of the indicators for each record
result_subset <- processed_mortality[, c("record_id", "drug_overdose_death", 
                                        "heroin_poisoning", "opioid_poisoning", 
                                        "fentanyl_poisoning", "stimulant_poisoning",
                                        "any_drug_poisoning")]
result_subset
```

## Understanding the Results

Let's examine each case in our sample dataset:

1. **Record 1**: Has underlying cause X42 (Accidental poisoning by narcotics) with multiple causes T40.1 (Poisoning by heroin) and F11.2 (Opioid dependence)
   - Identified as a drug overdose death with heroin poisoning and opioid poisoning

2. **Record 2**: Has underlying cause X44 (Accidental poisoning by unspecified drugs) with multiple causes T40.4 (Poisoning by synthetic narcotics) and T43.6 (Poisoning by psychostimulants)
   - Identified as a drug overdose death with fentanyl poisoning, opioid poisoning, and stimulant poisoning

3. **Record 3**: Has underlying cause I21.9 (Acute myocardial infarction) with multiple causes I10 (Essential hypertension) and E11.9 (Type 2 diabetes)
   - No drug-related indicators identified

4. **Record 4**: Has underlying cause Y14 (Poisoning by drugs of undetermined intent) with multiple cause T40.2 (Poisoning by other opioids)
   - Identified as a drug overdose death with opioid poisoning

5. **Record 5**: Has underlying cause V89.2 (Motor vehicle accident) with multiple causes S06.9 (Unspecified intracranial injury) and S27.0 (Traumatic pneumothorax)
   - No drug-related indicators identified

## CDC Definitions of Drug Overdose Deaths

According to the CDC and NCHS, drug overdose deaths are identified using the following ICD-10 underlying cause-of-death codes:

- X40–X44: Accidental poisoning by drugs
- X60–X64: Intentional self-poisoning by drugs
- X85: Assault by drugs
- Y10–Y14: Drug poisoning of undetermined intent

The specific drugs involved are identified using T-codes as multiple causes of death:

- T40.0: Opium
- T40.1: Heroin
- T40.2: Natural and semisynthetic opioids (e.g., oxycodone, hydrocodone)
- T40.3: Methadone
- T40.4: Synthetic opioids (e.g., fentanyl, tramadol)
- T40.5: Cocaine
- T40.7: Cannabis
- T40.8–T40.9: Hallucinogens
- T43.6: Psychostimulants with abuse potential (e.g., methamphetamine)

## Using Custom Variable Names

If your mortality dataset uses different variable names for the cause of death fields, you can specify them using the `icd_vars` parameter:

```{r}
# Create a sample with custom cause of death field names
sample_custom <- data.frame(
  record_id = 1:3,
  UCD = c("X42", "X44", "I21.9"),  # Underlying cause of death
  COD1 = c("T40.1", "T40.4", "I10"),  # Causes of death
  COD2 = c("F11.2", "T43.6", "E11.9"),
  COD3 = c(NA, NA, NA),
  stringsAsFactors = FALSE
)

# Process using custom variable names
processed_custom <- process_death_records_extended(sample_custom, 
                                                  icd_vars = c("COD1", "COD2", "COD3"))

# Check results
processed_custom[, c("record_id", "drug_overdose_death", "heroin_poisoning", 
                    "opioid_poisoning", "any_drug_poisoning")]
```

## Trends in Drug Overdose Deaths

The CDC has documented significant changes in drug overdose death patterns over time:

- From 1999 to 2018, drug overdose death rates increased from 6.1 per 100,000 to 20.7 per 100,000
- After a steady increase, there was a 4.1% decline in drug overdose deaths from 2017 to 2018
- Deaths involving synthetic opioids (mainly fentanyl) increased 10% from 2017 to 2018
- Deaths involving cocaine and psychostimulants (like methamphetamine) continued to increase

Using the `process_death_records_extended()` function allows researchers to track these patterns in their own mortality data.

## Summary

The `process_death_records_extended()` function provides a standardized method for identifying drug-related deaths in mortality data, following CDC definitions. This can be valuable for:

- Epidemiological studies of drug overdose mortality
- Surveillance of drug-related deaths
- Research on the patterns and trends of specific substances involved in overdose deaths
- Public health policy development and evaluation

The function handles large datasets efficiently through vectorized operations and can be adapted to different data structures through the `icd_vars` parameter.
